Morus: Biases in the stream

### 2018-05-09 morusAC.tex
split text into sections

\input{morusAC_00_Abstract}

\input{morusAC_01_Introduction}

\input{morusAC_02_Preliminaries}

\section{Symmetric Linear Combinations and \cipher{MiniMORUS}}
\label{sec:introminimorus}

To simplify the description of the attack, we assume all plaintext blocks are zero. However the same results would hold as long as the plaintext is fixed. Recall that the inner state of the cipher consists of five $4w$-bit registers $S_0,\dots,S_4$, each containing four $w$-bit words.

\subsection{Symmetric Linear Combinations}

We begin with a few observations about the \StateUpdate{} function. Besides XOR and AND operations, the \StateUpdate{} function uses two types of bit rotations:
\begin{enumerate}
\item \emph{word-wise} rotations perform a circular shift on each word within a register;
\item \emph{register-wise} rotations perform a circular shift on a whole register.
\end{enumerate}
The second type of rotation always shifts registers by a multiple of the word size $w$. This amounts to a (circular) permutation of the words within the register: for example, if a register contains the words $(A, B, C, D)$, and a register-wise rotation by $w$ bits to the left is performed, then the register now contains the words $(B, C, D, A)$.

Since we want to build a linear trail, we are interested in linear combinations of bits within a register.
\begin{definition}
Recall that $w$ denotes the word size in bits, and $4w$ is the size of a register. A linear combination of the form:
\[
S^t_{i,j(0)} \oplus S^t_{i,j(1)} \oplus \dots \oplus S^t_{i,j(k)}
\]
is said to be \emph{symmetric} iff the set of bits $S^t_{i,j(0)}, \dots, S^t_{i,j(k)}$ is left invariant by a circular shift by $w$ bits; that is, iff:
\[
\{j(i) : i\leq k\} = \{j(i) + w \text{\rm{} mod } 4w : i\leq k\}.
\]
\end{definition}
\emph{Example.} The following linear combination is symmetric for \cipher{MORUS640}, i.e. $w = 32$:
\begin{equation}
S^t_{0,0} \oplus S^t_{0,32} \oplus S^t_{0,64} \oplus S^t_{0,96}.
\label{eq:symmetric}
\end{equation}

This definition naturally extends to a linear combination across multiple registers, and also across ciphertext blocks.
The value of such a linear combination is unaffected by register-wise rotations, since those rotations always shift registers by a multiple of the word size.
On the other hand, since word-wise rotations always shift all four words within a register by the same amount, word-wise rotations preserve the symmetry property. Moreover, the XOR of two symmetric linear combinations is also symmetric.%; and the same holds for the AND operation (if we extend the symmetric property to non-linear combinations in the natural way).

This naturally leads to the idea of building a linear trail using only symmetric linear combinations, which is what we are going to do. As a result, the effect of register-wise rotations can be ignored. Moreover, since all linear combinations we consider are going to be symmetric, they can be described by truncating the linear combination to the first word of a register. Indeed, an equivalent way of saying a linear combination is symmetric, is that it involves the same bits in each word within a register. For example, in the case of (\ref{eq:symmetric}) above, the four bits involved are the first bit of each of the four words.

\subsection{\cipher{MiniMORUS}}

In fact, we can go further and consider a reduced version of \cipher{MORUS} where each register contains a single word instead of four. The \StateUpdate{} function is unchanged, except for the fact that register-wise rotations are removed: see Figure~\ref{fig:minimorus}. We call these reduced versions \cipher{MiniMORUS640} and \cipher{MiniMORUS1280}, for \cipher{MORUS640} and \cipher{MORUS1280} respectively. Since registers in \cipher{MiniMORUS} contain a single word, word-wise and register-wise rotations are the same operation; for simplicity we write $\lll$ for word-wise rotations.

Because the trail we are going to build is relatively complex, we will first describe it on \cipher{MiniMORUS}. We will then extend it to the full \cipher{MORUS} via the previous symmetry.

\begin{figure}[h]
  \substatesfalse
  % \substatesfalse to label state words and/or masks
  \centering
  \begin{tikzpicture}[xscale=1.0,yscale=1.5]%{{{
    \printstate
  \end{tikzpicture}%}}}
  \caption{\cipher{MiniMORUS} state update function.}
  \label{fig:minimorus}
\end{figure}

\section{Linear Trail for \cipher{MiniMORUS}}

\subsection{Overview of the Trail}

To build a linear trail on \cipher{MiniMORUS}, we combine the following five trail fragments $\alpha^t_i$, $\beta^t_i$, $\gamma^t_i$, $\delta^t_i$, $\varepsilon^t_i$, where the subscript $i$ denotes a bit position, and the superscript $t$ denotes a step number:
\begin{itemize}
    \item $\alpha^t_i$ approximates (one bit of) state word $S_0$ using the ciphertext;
    \item $\beta^t_i$ approximates $S_1$ using $S_0$ and the ciphertext;
    \item $\gamma^t_i$ approximates $S_4$ using two approximations of $S_1$ in consecutive steps;
    \item $\delta^t_i$ approximates $S_2$ using two approximations of $S_4$ in consecutive steps;
    \item $\varepsilon^t_i$ approximates $S_0$ using two approximations of $S_2$ in consecutive steps.
\end{itemize}

The trail fragments are depicted on \Cref{fig:minimorusfragments}. In all cases except $\alpha^t_i$, the trail fragment approximates a single AND gate by zero, which holds with probability 3/4, and hence the trail fragment has weight 1. In the case of $\alpha^t_i$, two AND gates are involved; however the two gates share an entry in common, and in both cases the other entry also has a linear contribution to the trail, which results in an overall contribution of the form:
\[
x \cdot y \oplus x \cdot z \oplus y \oplus z= (x \oplus 1) \cdot (y \oplus z).
\]
As a result, the trail fragment $\alpha^t_i$ also has a weight of 1. Another way of looking at this phenomenon is that the trail holds for two different approximations of the AND gates: the alternative approximation is depicted by a dashed line on \Cref{fig:minimorusfragments}.

\begin{figure}
  \substatesfalse
  % \substatesfalse to label state words and/or masks
  \centering
  \begin{subfigure}{.32\textwidth}
  \centering
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, alpha]
      (C) -- node[right] {$i$} (lll-1)
      (lll-1) -- (tlll-1) (lll-1) -- (xor-1) (xor-1) -- (xnd-1)
      (xnd-1) -- (and-1) (xor-1) -- (txor-1) (txor-1) -- (tanB0) (tanB0) -- (tanB0-|and0.east)
      (and0) -- (xnd0) (xnd0) -- (tlll-1) (xnd0) -- (xor0)
      (and-1.east|-tanA-1) -- (tanA-1) (tanA-1) -- (txor0) (txor0) -- (xor0)
      (xor0) -- (lll0) (lll0) -- (W40) node[below] {$i+b_0$}
      ;
    \draw[trail, alpha, dashed]
      (and-1.east|-tanB-1) -- (tanB-1) (tanB-1) -- (tanA0) (tanA0) -- (tanA0-|and0.east)
      ;
  \end{tikzpicture}%}}}
  \caption*{$\alpha^t_i$: weight 1 (not 2)} %$C^j_i, S^{j-1,0}_{i+5}$ ($w\!=\!1$)}
  \end{subfigure}
  \hfill
  \begin{subfigure}{.32\textwidth}
  \centering
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, beta]
      (C) -- node[right] {$i$} (lll-1)
      (lll-1) -- (tlll-1) (lll-1) -- (xor-1) (xor-1) -- (xnd-1)
      (tlll-1) -- (W-20) node[above] {$i$}
      (xor-1) -- (txor-1) (txor-1) -- (W-21) node[above] {$i$}
      (xnd-1) -- (and-1)
      (W40) node[below] {\phantom{$i$}}
      ;
  \end{tikzpicture}%}}}
  \caption*{$\beta^t_i$: weight 1} %$C^j_i, S^{j,0}_i, S^{j,1}_i$ ($w\!=\!0$)}
  \end{subfigure}
  \hfill
  \begin{subfigure}{.32\textwidth}
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, gamma]
      (W-21) node[above] {$i$} -- (xnd1) (xnd1) -- (xor1) (xor1) -- (lll1)
      (xnd1) -- (and1)
      (xor1) -- (txor1)
      (txor1) -- (W-24) node[above] {$i$}
      (lll1) -- (W41) node[below] {$i+b_1$}
      ;
  \end{tikzpicture}%}}}
  \caption*{$\gamma^t_i$: weight 1} %$S^{j,1}_i, S^{j,4}_i, S^{j+1,1}_{i-1}$ ($w\!=\!1$)}
  \end{subfigure}
  \bigskip

  \begin{subfigure}{.32\textwidth}
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, delta]
      (W-24) node[above] {$i$} -- (xnd4) (xnd4) -- (xor4) (xnd4) -- (and4)
      (xor4) -- (lll4) (xor4) -- (txor4) (txor4) -- (W42) node[below] {$i$}
      (lll4) -- (W44) node[below] {$i+b_4$}
      ;
  \end{tikzpicture}%}}}
  \caption*{$\delta^t_i$: weight 1} %$S^{j,4_i}, S^{j+1,2}_i, S^{j+1,4}_{i+13}$ ($w\!=\!1$)}
  \end{subfigure}
  \hfill
  \begin{subfigure}{.32\textwidth}
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, epsil]
      (W-22) node[above] {$i$} -- (xnd2) (xnd2) -- (xor2) (xnd2) -- (and2)
      (xor2) -- (lll2) (xor2) -- (txor2) (txor2) -- (W40) node[below] {$i$}
      (lll2) -- (W42) node[below] {$i+b_2$}
      ;
  \end{tikzpicture}%}}}
  \caption*{$\varepsilon^t_i$: weight 1} %$S^{j,2}_i, S^{j+1,0}_i, S^{j+1,2}_{i+7}$ ($w\!=\!1$)}
  \end{subfigure}
  \hfill
  \begin{subfigure}{.32\textwidth}
    \small
    \begin{align*}
      \intertext{\cipher{MiniMORUS-640}:}
\alpha^t_i:&& C^t_i &\to S^{t+1}_{0,i+5} \\
\beta^t_i:&& C^t_i, S^t_{0,i} &\to S^t_{1,i} \\
\gamma^t_i:&& S^t_{1,i}, S^{t+1}_{1,i+31} &\to S^t_{4,i} \\
\delta^t_i:&& S^t_{4,i}, S^{t+1}_{4,i+13} &\to S^{t+1}_{2,i} \\
\varepsilon^t_i:&& S^t_{2,i}, S^{t+1}_{2,i+7} &\to S^{t+1}_{0,i}.
      \intertext{\cipher{MiniMORUS-1280}:}
\alpha^t_i:&& C^t_i &\to S^{t+1}_{0,i+13} \\
\beta^t_i:&& C^t_i, S^t_{0,i} &\to S^t_{1,i} \\
\gamma^t_i:&& S^t_{1,i}, S^{t+1}_{1,i+46} &\to S^t_{4,i} \\
\delta^t_i:&& S^t_{4,i}, S^{t+1}_{4,i+4} &\to S^{t+1}_{2,i} \\
\varepsilon^t_i:&& S^t_{2,i}, S^{t+1}_{2,i+38} &\to S^{t+1}_{0,i}.
    \end{align*}
    \caption*{\cipher{MiniMORUS} instances}
  \end{subfigure}
  \caption{\cipher{MiniMORUS} linear trail fragments.}
  \label{fig:minimorusfragments}
\end{figure}

The way we are going to use each trail fragment may be summarized as follows, where in each case, elements to the left of the arrow $\to$ are used to approximate the element on the right of the arrow:
\begin{alignat*}{2}
\alpha^t_i:\quad&& C^t_i &\to S^{t+1}_{0,i+b_0} \\
\beta^t_i:\quad&& C^t_i, S^t_{0,i} &\to S^t_{1,i} \\
\gamma^t_i:\quad&& S^t_{1,i}, S^{t+1}_{1,i+b_1} &\to S^t_{4,i} \\
\delta^t_i:\quad&& S^t_{4,i}, S^{t+1}_{4,i+b_4} &\to S^{t+1}_{2,i} \\
\varepsilon^t_i:\quad&& S^t_{2,i}, S^{t+1}_{2,i+b_2} &\to S^{t+1}_{0,i}.
\end{alignat*}

In more detail, the idea is that by using $\alpha^t_i$, we are able to approximate a bit of $S_0$ using only a ciphertext bit. By combining $\alpha^t_i$ with $\beta^{t+1}_{i+b_0}$, we are then able to approximate a bit of $S_1$ (at step $t+1$) using only ciphertext bits from two consecutive steps. Likewise, $\gamma^t_i$ allows us to ``jump'' from $S_1$ to $S_4$, i.e. by combining $\alpha^t_i$ with $\beta^t_i$ and $\gamma^t_i$ with appropriate choices of parameters $t$ and $i$ for each, we are able to approximate one bit of $S_4$ using only ciphertext bits. Notice however that $\gamma^t_i$ requires approximating $S_1$ in two consecutive steps; and so the previous combination requires using $\alpha^t_i$ and $\beta^t_i$ \emph{twice} at different steps. In the same way, $\delta^t_i$ allows us to jump from $S_4$ to $S_2$; and $\varepsilon^t_i$ allows jumping from $S_2$ back to $S_0$. In the end, we are able to approximate a bit of $S_0$ using only ciphertext bits via the combination of all trail fragments $\alpha^t_i$, $\beta^t_i$, $\gamma^t_i$, $\delta^t_i$, and $\varepsilon^t_i$.

However, the same bit of $S_0$ can also be approximated directly by using $\alpha^t_i$ at the corresponding step. Thus that bit can be linearly approximated from two different sides: the first approximation uses a combination of all trail fragments, and involves successive approximations of all state registers (except $S_3$) spanning several encryption steps, as explained in the previous paragraph. The second approximation only involves using $\alpha^t_i$ at the final step reached by the previous trail. By XORing up these two approximations, we are left with only ciphertext bits, spanning five consecutive encryption steps.

Of course, the overall trail resulting from all of the previous combinations is quite complex, especially since $\gamma^t_i$, $\delta^t_i$, and $\varepsilon^t_i$ each require two copies of the preceding trail fragment in consecutive steps: that is, $\varepsilon^t_i$ requires two approximations of $S_2$, which requires using $\delta^t_i$ twice; and $\delta^t_i$ in turn requires using $\gamma^t_i$ twice, which itself requires using $\alpha^t_i$ and $\beta^t_i$ twice. Then $\alpha^t_i$ is used one final time to close the trail. One may naturally wonder if some components of this trail are in conflict. In particular, products of bits from registers $S_2$ and $S_3$ are approximated multiple times, by $\alpha^t_i$, $\beta^t_i$ and $\gamma^t_i$. To address this concern, and ensure that all approximations along the trail are in fact compatible, we now compute the full trail equation explicitly.

\subsection{Trail Equation}
\label{sec:minitraileq}

The equation corresponding to each of the five trail fragments $\alpha^t_i$, $\beta^t_i$, $\gamma^t_i$, $\delta^t_i$, $\varepsilon^t_i$ may be written explicitly as $A^t_i$, $B^t_i$, $C^t_i$, $D^t_i$, $E^t_i$ as follows. For each equation, we write on the left-hand side of the equality the biased linear combination used in the trail; and on the right-hand side, the remainder of the equation, which must have non-zero bias (in all cases the bias is $2^{-1}$).
\begin{alignat*}{2}
A^t_i:\quad&& C^t_i \oplus S^{t+1}_{0,i+b_0} &= S^t_{1,i} \oplus S^t_{3,i} \oplus S^t_{1,i} \cdot S^t_{2,i} \oplus S^t_{2,i} \cdot S^t_{3,i}\\
B^t_i:\quad&& C^t_i \oplus S^t_{0,i} \oplus S^t_{1,i} &= S^t_{2,i} \cdot S^t_{3,i} \\
C^t_i:\quad&& S^t_{1,i} \oplus S^{t+1}_{1,i+b_1} \oplus S^t_{4,i} &= S^t_{2,i} \cdot S^t_{3,i}\\
D^t_i:\quad&& S^t_{4,i} \oplus S^{t+1}_{4,i+b_4} \oplus S^{t+1}_{2,i} &= S^{t+1}_{0,i} \cdot S^{t+1}_{1,i}\\
E^t_i:\quad&& S^t_{2,i} \oplus S^{t+1}_{2,i+b_2} \oplus S^{t+1}_{0,i} &= S^t_{3,i} \cdot S^t_{4,i}
\end{alignat*}

From an algebraic point of view, building the full trail amounts to adding up copies of the previous equations for various choices of $t$ and $i$, so that eventually all $S^x_{y,z}$ terms on the left-hand side cancel out. Then we are left with only ciphertext terms on the left-hand side, while the right-hand side consists of a sum of biased expressions. By measuring the bias of the right-hand side expression, we will then be able to deduce the same bias on the linear combination of ciphertext bits on the left-hand side. We now set out to do so.

In order to build the equation for the full trail, we start with $E^2_0$:
\[
S^2_{2,0} \oplus S^{3}_{2,b_2} \oplus S^{3}_{0,0} = S^2_{3,0} \cdot S^2_{4,0}.
\]
In order to cancel the $S^{3}_{0,0}$ term on the left-hand side, we add in the equation $A^2_{-b_0}$ (where the sum of two equations of the form $a = b$ and $c = d$ is defined to be $a+c = b+d$). This yields:
\begin{align*}
&S^2_{2,0} \oplus S^3_{2,b_2} \oplus C^2_{-b_0}\\
=\; &S^2_{3,0} \cdot S^2_{4,0} \oplus S^2_{1,-b_0} \oplus S^2_{3,-b_0} \oplus S^2_{1,-b_0} \cdot S^2_{2,-b_0} \oplus S^2_{2,-b_0} \cdot S^2_{3,-b_0}.
\end{align*}
We then need to cancel two terms of the form $S^t_{2,i}$. To do this, we add in the equations $D^t_i$ for appropriate choices of $t$ and $i$. This replaces the two $S^t_{2,i}$ terms by four $S^t_{4,i}$ terms. By using equation $B^t_i$ four times, we can then replace these four $S^t_{4,i}$ terms by eight $S^t_{1,i}$ terms. By applying equation $B^t_i$ eight times, these eight $S^t_{1,i}$ terms can in turn be replaced by eight $S^t_{0,i}$ terms (and some ciphertext terms). Finally, applying $A^t_i$ eight times allows us to replace these eight $S^t_{0,i}$ terms by only ciphertext bits. Ultimately, for \cipher{MiniMORUS640}, this yields the equation:

\begin{flalign*}
&&&C^0_{27} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^2_{1} \oplus C^2_{13} \oplus C^2_{15}&&\\
&&\oplus\; &C^2_{27} \oplus C^2_{31} \oplus C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^4_{19}&&\\
&&=\; &S^0_{1,27} \oplus S^0_{3,27} \oplus S^0_{1,27} \cdot S^0_{2,27} \oplus S^0_{2,27} \cdot S^0_{3,27}&&\text{weight 1}\\
&&\oplus\; &S^1_{1,26} \oplus S^1_{3,26} \oplus S^1_{1,26} \cdot S^1_{2,26} \oplus S^1_{2,26} \cdot S^1_{3,26}&&\text{weight 1}\\
&&\oplus\; &S^1_{1,8} \oplus S^1_{3,8} \oplus S^1_{1,8} \cdot S^1_{2,8} \oplus S^1_{2,8} \cdot S^1_{3,8}&&\text{weight 1}\\
&&\oplus\; &S^2_{3,1} \oplus S^2_{1,1} \oplus S^2_{1,1} \cdot S^2_{2,1} \oplus S^2_{2,1} \cdot S^2_{3,1}&&\text{weight 1}\\
&&\oplus\; &S^2_{1,7} \oplus S^2_{3,7} \oplus S^2_{1,7} \cdot S^2_{2,7} \oplus S^2_{2,7} \cdot S^2_{3,7}&&\text{weight 1}\\
&&\oplus\; &S^2_{1,15} \oplus S^2_{3,15} \oplus S^2_{1,15} \cdot S^2_{2,15} \oplus S^2_{2,15} \cdot S^2_{3,15}&&\text{weight 1}\\
&&\oplus\; &S^2_{1,27} \oplus S^2_{3,27} \oplus S^2_{1,27} \cdot S^2_{2,27} \oplus S^2_{2,27} \cdot S^2_{3,27}&&\text{weight 1}\\
&&\oplus\; &S^1_{1,2} \oplus S^1_{3,2} \oplus S^1_{1,2} \cdot S^1_{2,2} \oplus S^1_{2,2} \cdot S^1_{3,2}&&\text{weight 1}\\
&&\oplus\; &S^3_{3,14} \oplus S^3_{1,14} \oplus S^3_{1,14} \cdot S^3_{2,14} \oplus S^3_{2,14} \cdot S^3_{3,14}&&\text{weight 1}\\
&&\oplus\; &S^2_{0,0} \cdot S^2_{1,0}&&\text{weight 1}\\
&&\oplus\; &S^2_{2,31} \cdot S^2_{3,31}&&\text{weight 1}\\
&&\oplus\; &S^2_{3,0} \cdot S^2_{4,0}&&\text{weight 1}\\
&&\oplus\; &S^3_{0,7} \cdot S^3_{1,7}&&\text{weight 1}\\
&&\oplus\; &S^3_{2,6} \cdot S^3_{3,6}&&\text{weight 1}\\
&&\oplus\; &S^3_{2,12} \cdot S^3_{3,12}&&\text{weight 1}\\
&&\oplus\; &S^4_{2,19} \cdot S^4_{3,19}.&&\text{weight 1}
\end{flalign*}

The equation for \cipher{MiniMORUS1280} is very similar, and is given in \Cref{sec:traileq}.

\subsection{Bias of the Trail}
\label{sec:minibias}

In the equation for \cipher{MiniMORUS640} from the previous section, each line on the right-hand side of the equality involves distinct $S^t_{i,j}$ terms, and each line has a weight of 1. By the piling-up lemma, it follows that if we assume distinct $S^t_{i,j}$ terms to be uniform and independent, then the expression on the right-hand side has a weight of 16. Hence the linear combination of ciphertext bits on the left-hand side has an expected bias of $2^{-16}$. The same holds for \cipher{MiniMORUS1280} (cf. \Cref{sec:traileq}).

That the bias is so high may be surprising: indeed, the full trail uses trail fragment $\varepsilon^t_i$ 1 time, $\delta^t_i$ 2 times, $\gamma^t_i$ 4 times, $\beta^t_i$ 8 times, and $\alpha^t_i$ 9 times. Since each trail fragment has a weight of 1, this would suggest that the total weight should be $1+2+4+8+9 = 24$ rather than 16. However, when combining trail fragments $\beta_i$ and $\gamma_i$, notice that the same AND is computed at the same step between registers $S_2$ and $S_3$ (equivalently, notice that the right-hand side of equations $B^t_i$ and $C^t_i$ is equal). In both cases it is approximated by zero. When XORing the corresponding equations, these two ANDs cancel each other, which saves two AND gates. Since $\gamma^t_i$ is used four times in the course of the full trail, this results in saving 8 AND gates overall, which explains why the final bias is $2^{-16}$ rather than $2^{-24}$.

\subsection{Experimental Verification}

To confirm that our analysis is correct, we ran experiments on an implementation of \cipher{MiniMORUS640} and \cipher{MiniMORUS1280}. In both cases, we measured the bias of two halves $\chi_1$ and $\chi_2$ of the full trail, as well as the bias of the full trail itself (denoted by $\chi$). Results are displayed on \Cref{tab:miniapproximations}.
While our analysis from the previous section predicts a bias of $2^{-16}$, experiments indicate a slightly higher bias of $2^{-15.5}$ for \cipher{MORUS640}. The discrepancy of $2^{-0.5}$ probably arises from the fact that register bits across different steps are not completely independent.

%\noindent
%2 approximations for $S^{2,2}_{0}$ of \cipher{MiniMORUS-640}:
%\begin{align*}
%  S^{2,2}_0 &= C^0_{27} \oplus C^1_{0, 8, 26} \oplus C^2_{7,13,31} \oplus C^3_{12} \tag{corr. $2^{-7}$} \\
%  S^{2,2}_0 &= C^1_{2} \oplus C^2_{1,7,15,27} \oplus C^3_{6,14,20} \oplus C^4_{19} \tag{corr. $2^{-9}$} \\
%\intertext{Combined ciphertext-only approximation:}
%  0 &= C^0_{27} \oplus C^1_{0, 2, 26, 8} \oplus C^2_{1,13,15,27,31} \oplus
%  C^3_{6,12,14,20} \oplus C^4_{19} \tag{corr. $\approx 2^{-15.5}$}
%\end{align*}
%\bigskip
%
%\noindent
%2 approximations for $S^{2,2}_{0}$ of \cipher{MiniMORUS-1280}:
%\begin{align*}
%  S^{2,2}_0 &= C^0_{51} \oplus C^1_{0, 33, 55} \oplus C^2_{4,37,46} \oplus C^3_{50} \tag{corr. $2^{-7}$} \\
%  S^{2,2}_0 &= C^1_{25} \oplus C^2_{7,29,38,51} \oplus C^3_{11,20,42} \oplus C^4_{24} \tag{corr. $2^{-9}$} \\
%\intertext{Combined ciphertext-only approximation:}
%  0 &= C^0_{51} \oplus C^1_{0, 25, 33, 55} \oplus C^2_{4,7,29,37,38,46,51} \oplus C^3_{11,20,42,50} \oplus C^4_{24} \tag{corr. $\approx 2^{-??}$}
%\end{align*}

\begin{table}[h!]
  \caption{Experimental verification of trail biases.}
  \label{tab:miniapproximations}
  \centerline{
  \begin{tabular}{@{}llSSS@{}}
    \toprule
    & & \multicolumn{3}{@{}c@{}}{Weight} \\
    \cmidrule{3-5}
    \multicolumn{2}{@{}l}{Approximations for \cipher{MiniMORUS640}}
             & {Exp.} & {Bool.} & {Meas.} \\
    \midrule
    $\chi_1$ & $S^{2,2}_0 = C^0_{27} \oplus C^1_{0, 8, 26} \oplus C^2_{7,13,31} \oplus C^3_{12}$
             & 7    & 7     & 7 \\
    $\chi_2$ & $S^{2,2}_0 = C^1_{2} \oplus C^2_{1,7,15,27} \oplus C^3_{6,14,20} \oplus C^4_{19}$
             & 9    & 9     & 9 \\
    $\chi$   & $0 = C^0_{27} \oplus C^1_{0, 2, 26, 8} \oplus C^2_{1,13,15,27,31} \oplus C^3_{6,12,14,20} \oplus C^4_{19}$
             & 16   & 16    & 15.5 \\
    \midrule
    \multicolumn{2}{@{}l}{Approximations for \cipher{MiniMORUS1280}} \\
    \midrule
    $\chi_1$ & $S^{2,2}_0 = C^0_{51} \oplus C^1_{0, 33, 55} \oplus C^2_{4,37,46} \oplus C^3_{50}$
             & 7    & 7     & 7 \\
    $\chi_2$ & $S^{2,2}_0 = C^1_{25} \oplus C^2_{7,29,38,51} \oplus C^3_{11,20,42} \oplus C^4_{24}$
             & 9    & 9     & 9 \\
    $\chi$   & $0 = C^0_{51} \oplus C^1_{0, 25, 33, 55} \oplus C^2_{4,7,29,37,38,46,51} \oplus C^3_{11,20,42,50} \oplus C^4_{24}$
             & 16   & 16    & 15.9 \\
    \bottomrule
  \end{tabular}
  }
\end{table}

\section{Trail for Full \cipher{MORUS}}

\subsection{Symmetrizing the Trail}
In the previous section, we have exhibited a linear trail on the reduced ciphers \cipher{MiniMORUS640} and \cipher{MiniMORUS1280}. We now turn to the full ciphers \cipher{MORUS640} and \cipher{MORUS1280}.
In order to build a trail on the full \cipher{MORUS}, we proceed exactly like we did for \cipher{MiniMORUS}, following the exact same path down to step and word rotation values, with one difference: in order to move from the one-word registers of \cipher{MiniMORUS} to the four-word registers of full \cipher{MORUS}, we make every term $S^t_{i,j}$ and $C^t_j$ symmetric, in the sense of \Cref{sec:introminimorus}.
That is, for every $S^t_{i,j}$ (resp. $C^t_j$) component in every trail fragment and every equation, we \emph{symmetrize} the term by adding in the terms $S^t_{i,j+w}$, $S^t_{i,j+2w}$, $S^t_{i,j+3w}$ (resp. $C^t_{j+w}$, $C^t_{j+2w}$, $C^t_{j+3w}$), where as usual $w$ denotes the word size. For example, if $w=32$ (for \cipher{MORUS640}), the term $S^3_{2,0}$ is symmetrized into:
\[
S^3_{2,0} \oplus S^3_{2,32} \oplus S^3_{2,64} \oplus S^3_{2,96}.
\]

Thus, translating the trail from one of the \cipher{MiniMORUS} ciphers to the corresponding full \cipher{MORUS} cipher amounts to making every linear combination symmetric---indeed, that was the point of introducing \cipher{MiniMORUS} in the first place.
Concretely, in order to build the full trail equation for \cipher{MORUS}, we write symmetric versions of equations $A^t_i$, $B^t_i$, $C^t_i$, $D^t_i$, $E^t_i$ from \Cref{sec:minitraileq}, and then combine them in exactly the same way as before.
In this way, the biased linear combination on \cipher{MiniMORUS640} given in \Cref{sec:minitraileq}, namely:
\begin{align*}
&C^0_{27} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^2_{1} \oplus C^2_{13} \oplus C^2_{15}\\
\oplus\; &C^2_{27} \oplus C^2_{31} \oplus C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^4_{19}
\end{align*}
ultimately yields the following biased symmetrized linear combination on the full \cipher{MORUS640}:
\begin{align*}
&C^0_{27} \oplus C^0_{59} \oplus C^0_{91} \oplus C^0_{123} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^1_{32} \oplus C^1_{34}\\
\oplus\; & C^1_{40} \oplus C^1_{58} \oplus C^1_{64} \oplus C^1_{66} \oplus C^1_{72} \oplus C^1_{90} \oplus C^1_{96} \oplus C^1_{98} \oplus C^1_{104} \oplus C^1_{122}\\
\oplus\; & C^2_{1} \oplus C^2_{13} \oplus C^2_{15} \oplus C^2_{27} \oplus C^2_{31} \oplus C^2_{33} \oplus C^2_{45} \oplus C^2_{47} \oplus C^2_{59} \oplus C^2_{63}\\
\oplus\; & C^2_{65} \oplus C^2_{77} \oplus C^2_{79} \oplus C^2_{91} \oplus C^2_{95} \oplus C^2_{97} \oplus C^2_{109} \oplus C^2_{111} \oplus C^2_{123} \oplus C^2_{127}\\
\oplus\; & C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^3_{38} \oplus C^3_{44} \oplus C^3_{46} \oplus C^3_{52} \oplus C^3_{70} \oplus C^3_{76}\\
\oplus\; & C^3_{78} \oplus C^3_{84} \oplus C^3_{102} \oplus C^3_{108} \oplus C^3_{110} \oplus C^3_{116} \oplus C^4_{19} \oplus C^4_{51} \oplus C^4_{83} \oplus C^4_{115}.
\end{align*}
We refer the reader to \Cref{sec:traileq} for the corresponding linear combination on \cipher{MORUS1280}.

\subsection{Bias of the Full Trail}

The symmetrized trail on full \cipher{MORUS} may be intuitively understood as consisting of four copies of the original trail on \cipher{MiniMORUS}. Indeed, the only difference between full \cipher{MORUS} (for either version of \cipher{MORUS}) and four independent copies of \cipher{MiniMORUS} comes from register-wise rotations, which permute words within a register. But as observed in \Cref{sec:introminimorus}, register-wise rotations leave symmetric linear combinations invariant; and so, insofar as we only ever use symmetric linear combinations on all registers along the trail, register-wise rotations have no effect.

Following the previous intuition, one may expect that the weight of the symmetrized trail should simply be four times the weight of the corresponding \cipher{MiniMORUS} trail, namely 64 for both \cipher{MORUS640} and \cipher{MORUS1280}. However, reality is a little more complex, as the symmetrized trail does not exactly behave as four copies of the original trail when one considers nonlinear terms.

To understand why that might be the case, assume a nonlinear term $S^0_{2,0} \cdot S^0_{3,0}$ arises from some part of the trail, and another term $S^0_{2,0} \cdot S^0_{3,w}$ arises from a different part of the trail (where $w$ denotes the word size). Then when we XOR the various trail fragments together, in \cipher{MiniMORUS} these two terms are actually equal and will cancel out, since register-wise rotations by multiples of $w$ bits are ignored. However in the real \cipher{MORUS} these terms are of course distinct and do not cancel each other.

In the actual trail for (either version of) full \cipher{MORUS}, this exact situation occurs when combining trail fragments $\beta^t_i$ and $\gamma^t_i$. Indeed, $\beta^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t_{3,i}$, while $\gamma^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t_{3,i-w}$ (cf. \Cref{fig:trailcollision}). While in \cipher{MiniMORUS}, these terms cancel out, in the full \cipher{MORUS}, when adding all symmetric copies of the trail, we end up with the sum:
\begin{align}
&S^t_{2,i} \cdot S^t_{3,i} \oplus S^t_{3,i} \cdot S^t_{2,i+w}
\oplus S^t_{2,i+w} \cdot S^t_{3,i+w} \oplus S^t_{3,i+w} \cdot S^t_{2,i+2w}\notag\\
\oplus\; &S^t_{2,i+2w} \cdot S^t_{3,i+2w} \oplus S^t_{3,i+2w} \cdot S^t_{2,i+3w}
\oplus S^t_{2,i+3w} \cdot S^t_{3,i+3w} \oplus S^t_{3,i+3w} \cdot S^t_{2,i}.\label{eq:8circle}
\end{align}
It may be observed that the products occurring in the equation above involve eight terms forming a ring. The weight of this expression can be computed by brute force, and is equal to $3$.

>>>>

\input{morusAC_00_Abstract}
\input{morusAC_01_Introduction}
\input{morusAC_02_Preliminaries}
\input{morusAC_03_MiniMorus}
\input{morusAC_04_Mini_trails}
\input{morusAC_05_Full_trails}
\input{morusAC_06_Discussion}
\input{morusAC_07_IniFin}
